---
title: 'Reproducible Research: Peer Assessment 1'
output:
  md_document: default
  html_document: default
keep_md: yes
---


## Loading and preprocessing the data

Load library

```{r library, echo=TRUE}
library(dplyr)
library(ggplot2)
library(lubridate)

```

Read the data and assign it's content to "activity"

```{r readdata, echo=TRUE}
activity <- read.table(unz("activity.zip", "activity.csv"), header=T, sep=",")
print("See preview of loaded data")
head (activity)

```

Process or transform the date field into the format "Year-Month-Day

```{r datetransform, echo=TRUE}

activity$date <- as.Date(activity$date, format = "%Y-%m-%d")

```

## What is mean total number of steps taken per day?

Compute total number of steps taken each day

```{r totalsteps, echo=TRUE}
totalsteps <- activity %>% group_by(date) %>% summarise(total.steps = sum(steps, na.rm = TRUE))

head(tibble(totalsteps))

```

Histogram of the number of steps taken each day

```{r histview, echo=TRUE}

g <- ggplot(totalsteps, aes(x = date, y = total.steps))
g <- g + geom_histogram(stat = "identity", colour = " black", fill = "green")
g <- g + labs(title = "Histogram of the total number of steps taken each day", x = "Date", y = "Total steps")
print(g)

```


Compute the mean of number of steps taken each day

```{r mean, echo=TRUE}

mean <- mean(totalsteps$total.steps, na.rm = TRUE)
mean

```

Compute the median of number of steps taken each day
```{r median, echo=TRUE}

median <- median(totalsteps$total.steps, na.rm = TRUE)
median

```
## Average daily activity pattern?

Compute daily average steps 

```{r activitypattern, echo=TRUE}

averagesteps <- activity %>% group_by(interval) %>% summarise(avg.steps = mean(steps, na.rm = TRUE))

head(tibble(averagesteps))

```
Plot time series

```{r timeseries, echo=TRUE}

g <- ggplot(averagesteps, aes(x = interval, y = avg.steps))
g <- g + geom_line(stat = "identity", col = "blue")
g <- g + labs(title = "Time series plot of the average number of steps taken", x = "Intrval", y = "Average steps")

print(g)
```

Compute the interval with the maximum number of steps on average

```{r maxmeanstep, echo=TRUE}

intwthmaxsteps <- averagesteps$interval[averagesteps$avg.steps == max(averagesteps$avg.steps)]
intwthmaxsteps

```
The maximum average steps coincides with the 835th 5 minute interval

## Imputing missing values

Total number of missing values in data set

```{r missingvaluecount, echo=TRUE}

missingvalue <- activity[is.na(activity) == TRUE]
length(missingvalue)
```

## Method of of filling in missing values

The method used is to group all data according to their 5-minute intervals. The means of the 5-minute intervals are computed without NAs. The respective mean values are assigned to all locations where NA is true.

```{r imputing, echo = TRUE}
fills <-  function(x){ 
                 x$steps[is.na(x$steps)] <- ave(x$steps, x$interval, FUN = function(z) 
                 mean(z, na.rm = TRUE))[c(which(is.na(x$steps)))]
return(x)
}
activity_fld <- fills(activity)
head(tibble(activity_fld))

```

Histogram view of the number of steps taken each day over filled data

```{r histview2, echo=TRUE}

totalsteps2 <- activity_fld %>% group_by(date) %>% summarise(total.steps = sum(steps))
head(tibble(totalsteps2))

## Plotting total steps
g <- ggplot(totalsteps2, aes(x = date, y = total.steps))
g <- g + geom_histogram(stat = "identity", colour = " black", fill = "green")
g <- g + labs(title = "Histogram of the total number of steps taken each day after missing values are imputed", x = "Date", y = "Total steps")
print(g)

```

Compute the new mean of number of steps per day

```{r mean2, echo=TRUE}

mean <- mean(totalsteps2$total.steps)

mean

```

Compute the new median of number of steps per day

```{r median2, echo=TRUE}

median <- median(totalsteps2$total.steps)
median

```

The new mean and median are different from the original ones computed by removing NAs.

## Are there differences in activity patterns between weekdays and weekends?

The activity patterns between the weekday and weekends are different. See figure below:

```{r wk_days_end, echo=TRUE}

activity_fld$wk_days <- weekdays(as.Date(activity_fld$date))

i = 1

for (i in 1:length(activity_fld$wk_days)){
    if(activity_fld$wk_days[i] == "Saturday"|activity_fld$wk_days[i] == "Sunday"){
        activity_fld$wk_grp[i] <- "Weekend"
    } else {
      activity_fld$wk_grp[i] <- "Weekday"
      }
}

averagesteps2 <- activity_fld %>% group_by(interval, wk_grp) %>% summarise(avg.steps = mean(steps, na.rm = TRUE))

head(tibble(averagesteps2))

library(lattice)
g <- ggplot(averagesteps2, aes(x = interval, y = avg.steps))
g <- g + geom_line(stat = "identity", col = "blue")
g <- g + facet_wrap(~ wk_grp, ncol = 1)
g <- g + labs(title = "Comparing the average number of steps taken per 5-minute interval", x = "Interval", y ="Average steps")
print(g)

```

